---
title: "Analyzing the Relationship Between Life Expectancy and GDP per Capita"
author: "Shamima Haque"
date: "2023-10-18"
output:
  html_document: default
  word_document: default
editor_options: 
  markdown: 
    wrap: 72
---

#Introduction/Background

In the realm of data analytics, understanding the intricate dynamics
between life expectancy and Gross Domestic Product (GDP) per capita
stands as a pivotal endeavor. Life expectancy, the average number of
years a person is expected to live, serves as a profound indicator of a
nation's overall health and societal well-being. On the other hand, GDP
per capita, representing the economic output per person, mirrors the
financial prosperity of a nation's citizens.

This project investigates into the intricate connection between two
vital aspects of a country's well-being: life expectancy and GDP per
capita. Inspired by the impactful Gapminder project, co-founded by the
esteemed Hans Rosling, a leading figure in statistics and public health,
this venture leverages the extensive and globally diverse Gapminder
dataset. Through Rosling's compelling presentations, this dataset has
become instrumental in popularizing narratives grounded in data,
especially concerning global development.

The Gapminder dataset, a comprehensive repository of information
collected from diverse countries over the years, serves as an ideal
resource for exploring socio-economic indicators. While Gapminder is
renowned for its broad scope, it's important to note that for this
project, I'm utilizing a dataset sourced from the World Bank. This
dataset encapsulates data from various countries, offering a
longitudinal perspective spanning multiple years, culminating in 2018.

In delving into the intricate factors influencing life expectancy, such
as happiness, pollution, terrorism, and diseases, my primary focus has
been on a specific aspect: the correlation between life expectancy and
the GDP per capita of each country. The results underscore a notable
trend -- countries with a higher GDP per capita generally exhibit a
superior life expectancy for their citizens, surpassing outcomes
observed in countries with a lower GDP per capita.

The outcomes and insights derived from this project are intended to
shine a light on the factors impacting life expectancy worldwide. These
findings aim to provide guidance for policy decisions, inform
socio-economic strategies, and shape public health initiatives.
Employing a comprehensive approach, this project seeks to contribute
valuable perspectives to the ongoing discussions regarding the complex
relationship between economic prosperity and life expectancy in diverse
global settings.

# Problem Statement

The primary goal of this project is to unravel how life expectancy
correlates with the economic prosperity of nations. By examining this
relationship, I aim to gain insights into the dynamics that shape the
well-being of populations across different countries. Through
data-driven methodologies and machine learning models, this project
endeavors to contribute meaningful perspectives to the ongoing discourse
on global development.

#Data collection and pre-processing

In this project datasets sourced from the World Bank for the analysis.
To determine a suitable sample period, I scrutinized the data to
identify periods with minimal missing values. Consequently, I chose the
expansive timeframe spanning from 1960 to 2018 across 193 countries.
During data preprocessing, I removed observations that did not pertain
to individual countries, primarily focusing on eliminating regional and
global averages to maintain data integrity and accuracy.

The initial step in analyzing the collected dataset involved meticulous
pre-processing. Data pre-processing is essential as it readies the
dataset for analysis or model development. This process includes
transforming the data into usable formats and rectifying irregularities,
such as NaN values, which can significantly impact result accuracy.

The dataset extracted for this study was in the desired CSV format.
However, it did contain missing or unidentified values. Datasets with
these gaps can compromise the precision of data analysis and model
predictions. Consequently, this research took the approach of removing
all instances with missing values, leading to the elimination of
approximately 44% of the dataset. Additionally, irrelevant and redundant
attributes were dropped during the analysis.

```{r}

install.packages("ggplot2")
library(ggplot2)
install.packages("readxl")

# Load the readxl package
library(readxl)

data <- read.csv("Life Expectancy vs GDP.csv", header = TRUE)

head(data)  # To see the first few rows of the dataset
tail(data)
summary(data)

```

```{r}

# Check the current column names
print(colnames(data))


# Change the column name from 'GDP.par Cap' to 'GDP'
colnames(data)[colnames(data) == "GDP.per.capita"] <- "gdp_per_capita"

# Change the column name from 'GDP.par Cap' to 'GDP'
colnames(data)[colnames(data) == "Life.expectancy"] <- "LifeExp"

colnames(data)[colnames(data) == "Population..historical.estimates."] <- "pop"


colnames(data)[colnames(data) == "Entity"] <- "country"
colnames(data)[colnames(data) == "Continent"] <- "continent"
colnames(data)[colnames(data) == "Year"] <- "year"

# Verify the changes
print(colnames(data))

# Remove missing values (if any)
data <- na.omit(data)

head(data)
tail(data)

```

## Data Splitting:

Before building any model it's essential to preprocess the data. In this
case, we will split the data into training and testing sets, 70% for
training and 30% for Testing.

```{r}

# Select only the necessary columns (GDP per capita, life expectancy, and year)
data_set <- data %>% select(gdp_per_capita, LifeExp, year)

# Split the data into training and testing sets
set.seed(123) # for reproducibility
train_index <- createDataPartition(data_set$LifeExp, p = 0.7, 
                                   list = FALSE, 
                                   times = 1)

train_data <- data_set[train_index, ]
test_data <- data_set[-train_index, ]


```

# Methods

Visualization plays a crucial role in this exploration, illustrating the
temporal evolution of life expectancy across different countries
throughout the years. Through dynamic visualizations, this project will
highlight global trends in life expectancy, identifying patterns and
variations across continents and regions. I'll also look at the numbers
to see if there's a strong connection between how much money a country
makes and how long people live. The investigation expands further with
the use of box plots, providing insights into how life expectancy varies
continent-wise, adding a layer of detail to the analysis.

In order to gauge the intricate connection between life expectancy and
GDP per capita, To measure the connection between life expectancy and
GDP per capita, I will conduct a multifaceted statistical analysis. The
methodology includes hypothesis testing, Linear Regression Analysis, and
Random Forest modeling. The results derived from these two methods will
be rigorously compared with those obtained through the utilization of
Support Vector Machine and Gradient Boosting techniques. Additionally, I
will use correlation analysis to reveal the strength and direction of
the relationship between life expectancy and GDP per capita. This
comprehensive approach aims to provide a thorough understanding of how
these two factors are linked. Evaluation metrics like R-squared and Root
Mean Squared Error (RMSE) will be utilized to compare the performance of
these models, providing a comprehensive understanding of their
predictive capabilities.

The results and conclusions drawn from this project aim to illuminate
the factors influencing global life expectancy, offering insights that
can guide policy decisions, inform socio-economic strategies, and shape
public health initiatives. Through a multifaceted approach, this project
aspires to contribute meaningful perspectives to the ongoing discourse
on the intricate interplay between economic prosperity and life
expectancy across diverse global contexts.

# Choropleth Map

A choropleth map is a type of thematic map that uses colors or shading
to represent statistical data across geographic regions or areas. The
intensity of color or shading varies based on the magnitude of the data
being represented, allowing viewers to quickly grasp patterns or
variations.

```{r}
# Install the required packages
install.packages(c("plotly", "viridis"))

# Load the libraries
library(plotly)
library(viridis)

# Filter the data for the year 2018
filtered_data <- data[data$year == 2018, ]

# Create a choropleth map
choropleth_map <- plot_geo(
  data = filtered_data,
  locationmode = "country names",
  locations = ~Country,  # Replace with the appropriate column name
  z = ~LifeExp,
  text = ~Country
) %>%
  add_trace(
    z = ~LifeExp,
    colorbar = list(title = 'Life Expectancy'),
    colorscale = viridis(100, option = "D")
  ) %>%
  layout(
    title = 'Life Expectancy Choropleth Map Year of 2018',
    geo = list(
      projection = list(type = 'natural earth')
    )
  )

# Show the map
choropleth_map

```


```{r}

install.packages(c("plotly", "viridis"))

library(plotly)
library(viridis)

filtered_data <- data[data$year == 2018, ]

choropleth_map <- plot_geo(
  data = filtered_data,
  locationmode = "country names",
  locations = ~Country,
  z = ~LifeExp,
  text = ~Country
)

choropleth_map <- add_trace(
  choropleth_map,
  z = ~LifeExp,
  colorbar = list(title = 'Life Expectancy'),
  colorscale = viridis(100, option = "D")
)

choropleth_map <- layout(
  choropleth_map,
  title = 'Life Expectancy Choropleth Map Year of 2018',
  geo = list(
    projection = list(type = 'natural earth')
  )
)

choropleth_map

```

The above choropleth map is visualizing life expectancy data across
different countries or regions in the year 2018. Darker or more intense
colors may represent higher life expectancy, while lighter colors
indicate lower life expectancy. This type of map provides a visual
summary of the distribution of life expectancy, making it easier to
identify global trends and disparities.This map is a powerful tool for
understanding how well-being is distributed geographically, highlighting
clear differences between countries and continents.

##Scatter Plot for Life Expectancy vs GDP per Capita

A scatter plot is useful for identifying trends, correlations, or
patterns in the data. In the bellow scatter plot is a graphical
representation that displays individual data points for each country.

```{r}
install.packages("plotly")
library(plotly)
# Filter the data for the year 2018
data_2018 <- subset(data, year == 2018)

# Create a scatter plot with color based on the continent for the year 2018
fig <- plot_ly(data_2018, x = ~gdp_per_capita, y = ~LifeExp, color = ~continent,
               size = ~pop,  # Use the 'pop' column for bubble size
               colors = 'Set1', mode = "markers",
               text = ~paste("Country: ", Country, "<br>Continent: ", continent, "<br>Population: ", pop)) %>%
       layout(title = "Scatter Plot: GDP per capita vs Life Expectancy (2018)",
              xaxis = list(title = "GDP per capita"),
              yaxis = list(title = "Life Expectancy"))

# Show the plot
fig

```

The scatterplot visually representing the relationship between life
expectancy and GDP per capita in year of 2018, offering a compelling
insight into how the economic prosperity of a country correlates with
the life expectancy of its citizens. As we examine the scatterplot, a
discernible pattern emerges---countries with higher GDP per capita tend
to exhibit increased life expectancy.

Each point on the plot corresponds to a specific country, with the
x-axis representing the GDP per capita and the y-axis representing life
expectancy. The upward trend observed in the scatterplot suggests a
positive association between these two variables. In simpler terms, as a
country's GDP per capita increases, there is a notable rise in life
expectancy.

This relationship underscores the potential impact of economic
well-being on the overall health and longevity of a population. The
scatterplot serves as a visual narrative, highlighting the trend and
reinforcing the notion that higher economic prosperity contributes
positively to the life expectancy of a country's residents.

# Box Plots

A box plot showcasing life expectancy data by continent provides a
concise and insightful visualization of the distribution and central
tendencies of life expectancy across different regions. Each box
represents a continent, displaying the interquartile range (IQR),
median, and potential outliers.

```{r}
par(mfrow=c(1,2))

# Box plot of Life Expectancy by Continent
ggplot(data, aes(x = continent, y = LifeExp, fill = continent)) +
  geom_boxplot() +
  ggtitle("Life Expectancy by Continent") +
  xlab("Continent") +
  ylab("Life Expectancy")

# Box plot of GPD per Capita by Continent
ggplot(data, aes(x = continent, y = gdp_per_capita, fill = continent)) +
  geom_boxplot() +
  ggtitle("GDP per capita by Continent") +
  xlab("Continent") +
  ylab("GDP per capita")


```

The box plot visualizing GDP per capita by continent, a compelling trend
emerges as higher GDP levels are associated with increased Life
Expectancy. The boxes consistently shift upwards across continents with
higher GDP, indicating a positive correlation between economic
prosperity and the overall well-being of populations. This observation
suggests that countries with higher GDP per capita tend to exhibit not
only stronger economic performance but also elevated life expectancies,
providing valuable insights into the interconnected dynamics of economic
development and public health on a global scale.

# Hypothesis Test

Hypothesis testing for correlation involves testing whether the observed
correlation coefficient is significantly different from zero.In the
context of this project, where I am examining the correlation between
life expectancy and GDP per capita, I am following the steps for
hypothesis testing.

$$
\begin{align*}
{Hypothesis\ Test:} \\
\begin{align*}
& H_0: \text{There is no significant correlation between life expectancy and GDP per capita (\(\rho = 0\)).} \\
& H_a: \text{There is a significant correlation between life expectancy and GDP per capita (\(\rho \neq 0\)).}
\end{align*}
$$

$$
\begin{align*}
{Interpret\ Results:}
\begin{align*}
& \text{If p-value} < \alpha, \text{reject the null hypothesis.} \\
& \text{If p-value} \geq \alpha, \text{fail to reject the null hypothesis.}
\end{align*}
$$

```{r}

# Calculate Correlation Coefficient
correlation_coefficient <- cor(data$LifeExp, data$gdp_per_capita)

# Conduct Hypothesis Test
cor_test_result <- cor.test(data$LifeExp, data$gdp_per_capita)

# Interpret Results
cat("Observed Correlation Coefficient:", correlation_coefficient, "\n")
cat("p-value:", cor_test_result$p.value, "\n")

# Check if the p-value is less than the significance level (e.g., 0.05)
if (cor_test_result$p.value < 0.05) {
  cat("Reject the null hypothesis. There is a significant correlation.\n")
} else {
  cat("Fail to reject the null hypothesis. There is not enough evidence of a significant correlation.\n")
}

```

The observed correlation coefficient of 0.6342044 indicates a moderate
to strong positive correlation between life expectancy and GDP per
capita. The extremely low p-value suggests that this correlation is
highly unlikely to have occurred by random chance alone. Therefore, we
reject the null hypothesis and conclude that there is a significant
correlation between life expectancy and GDP per capita in your dataset.

# Correlation Matrix:

A correlation matrix is a statistical table that displays the
correlation coefficients between several variables. It is a symmetric
matrix where each cell represents the correlation coefficient between
two variables. Correlation coefficients quantify the strength and
direction of a linear relationship between two variables.

The correlation coefficient is a numerical value ranging from -1 to 1.
Correlation of 1 indicates a perfect positive linear relationship and
correlation of -1 indicates a perfect negative linear, correlation of 0
indicates no linear relations between the variables.

```{r}
# Visualize the correlation matrix as a heatmap using corrplot
install.packages("corrplot")  # Install corrplot package if you haven't already
library(corrplot)

# Customize and display the correlation heatmap
corrplot(
  correlation_matrix,                  # Data: correlation matrix
  method = "color",                    # Color representation of correlation values
  type = "upper",                      # Display upper triangle only (to avoid redundancy)
  tl.col = "gray",                    # Label color
  tl.srt = 45,                         # Label rotation angle
  title = "Correlation Coefficient\n(Life Expectancy, GDP per Capita, Population)",  # Plot title
  addCoef.col = "gray",               # Color of correlation coefficients
  tl.pos = "lt",                       # Label position: left and top
  tl.cex = 0.8                         # Label text size
)

# Calculate the correlation matrix between Population, Life Expectancy, and GDP per Capita
correlation_matrix <- cor(data[c("LifeExp", "gdp_per_capita", "pop")])

# Print the correlation matrix
print("Correlation Matrix:")
print(correlation_matrix)


```

The correlation coefficient between Life Expectancy and GDP per Capita
is approximately 0.60. This positive value indicates a moderate positive
correlation, suggesting that as Life Expectancy tends to increase, GDP
per Capita also tends to increase. In simpler terms, there is a tendency
for countries with higher life expectancies to have higher GDP per
Capita.

# Model Building and Prediction

## Linnear Regression Model

Linear Regression is a machine learning algorithm used to model the
relationship between a dependent variable and one or more independent
variables. It assumes a linear relationship between the variables,
aiming to find the best-fitting line that minimizes the sum of squared
differences between observed and predicted values.

\$\$

Linear Regression Equation: \\ Y = \beta\_0 + \beta\_1 X_1 + \beta\_2
X_2 + \ldots + \beta\_n X_n + \varepsilon \\ $$

$$ \\begin{align\*} \\

In this equation: \\ Y: \text{Dependent variable}\\

\beta\_0: \text{Y-intercept, the constant term}\\

\beta\_1, \beta\_2, \ldots, \beta\_n:
\text{Coefficients of the independent variables } (X_1, X_2, \ldots,
X_n)\\

X_1, X_2, \ldots, X_n: \text{Independent variables}\\

\varepsilon:
\text{Error term, representing unobserved factors affecting the dependent variable.}

\$\$

Linear regression model to predict life expectancy:

```{r}

# Build a linear regression model
lm_model <- lm(LifeExp ~ gdp_per_capita, data = train_data)

summary(lm_model)

```

In the linear regression model suggests a significant positive
relationship between GDP per capita and life expectancy. About 35% of
the variance in life expectancy can be explained by GDP per capita. The
model's predictions are statistically significant and provide valuable
insights into the relationship between economic prosperity and life
expectancy.

# Linear Regression Scatter Plot

```{r}

# Load the necessary libraries
library(plotly)

# Create a scatter plot of actual vs predicted values
scatter_plot <- plot_ly(data = test_data, 
                        x = ~LifeExp, 
                        y = ~predictions, 
                        text = ~paste("Actual: ", LifeExp, "<br>Predicted: ", predictions)) %>%
  add_trace(type = "scatter", mode = "markers", marker = list(size = 7, opacity = 0.9)) %>%
  add_trace(x = ~LifeExp, y = ~LifeExp, mode = "lines", 
            line = list(color = 'red', width = 2, dash = 'dash'),
            text = 'Ideal Prediction Line') %>%
  layout(title = "Actual vs. Predicted Life Expectancy",
         xaxis = list(title = "Actual Life Expectancy"),
         yaxis = list(title = "Predicted Life Expectancy"),
         hovermode = "closest")

# Show the plot
scatter_plot

```

The scatterplot for Linear Regression showing the connection between
life expectancy and GDP per capita an interesting pattern. Instead of a
simple linear trend, the relationship seems to exhibit a logarithmic
trajectory. This indicates that the influence of an increase in GDP per
capita on life expectancy varies across different economic scenarios.

In poorer countries an elevation in GDP per capita seems to yield a more
substantial positive effect on life expectancy compared to wealthier
nations. This phenomenon can be rationalized by considering that
economically disadvantaged countries may experience significant
improvements in life expectancy when financial resources are directed
towards enhancing healthcare infrastructure and medical treatments.

#Random Forest Regressor model

In order to explore a more advanced approach to predict life expectancy
based on GDP per capita. I'm using a Random Forest regressor.

Random Forest is a powerful ensemble learning algorithm used for both
classification and regression tasks. It was introduced by Leo Breiman in
2001. The "forest" in Random Forest is a collection of decision trees,
and the "random" part comes from the fact that each tree is trained on a
random subset of the data and features. $$
\text{Regression:}\\ \hat{y}_{\text{RF}} = \frac{1}{n} \sum_{i=1}^{n} \hat{y}_i 
$$

\$\$ \text{Classification}: \\

\hat{y}\_{\text{RF}} = \text{mode}(\hat{y}\_1, \hat{y}\_2, \ldots,
\hat{y}\_n) \$\$ Random Forest is widely used in various fields due to
its versatility and effectiveness in producing high-quality predictions.

```{r}

# Load libraries
library(caret)
library(randomForest)
library(dplyr)

# Build a Random Forest Regressor model
rf_model <- randomForest(LifeExp ~ gdp_per_capita, data = train_data, ntree = 100)

# Make predictions on the test set
test_data$predictions_rf <- predict(rf_model, newdata = test_data)

# Calculate R-squared score for Random Forest
rf_r_squared <- 1 - sum((test_data$LifeExp - test_data$predictions_rf)^2) / sum((test_data$LifeExp - mean(test_data$LifeExp))^2)
print(paste("Random Forest R-squared Score: ", round(rf_r_squared, 2)))

# Calculate RMSE for Random Forest Regressor

rmse_rf <- sqrt(mean((test_data$predictions_rf - test_data$LifeExp)^2))
rmse_rf
rf_model 

```

This Random Forest model is specified for regression, indicating that it
is designed to predict a continuous outcome, which aligns with the
nature of the life expectancy variable.

The Random Forest consists of an ensemble of decision trees. In this
case, there are 100 trees in the forest. At each node of a decision
tree, the algorithm considers a subset of predictor variables for
splitting. Here only 1 variable is tried at each split. This can
contribute to the diversity of the trees in the ensemble. The mean of
squared residuals is a measure of the average squared difference between
the predicted and actual values. In this context, it is 70.18432. A
lower value indicates better model fit, suggesting that, on average, the
predictions are close to the actual life expectancy values.

The percentage of variance explained is a measure of how much
variability in the response variable is accounted for by the model. In
this case, the model explains 56.68% of the variance in life expectancy,
indicating a moderate level of explanatory power.

Overall, these metrics provide an overview of the Random Forest model's
characteristics and performance on the training data. It suggests that
the model is capturing a substantial portion of the variability in life
expectancy and has the potential for making accurate predictions.
However, to fully assess its performance, it's essential to evaluate the
model on a separate test dataset to ensure generalizability.

# Random Forest Scatter Plot

The output of the Random Forest model indicates promising
characteristics, with a moderate percentage of variance explained
(56.68%) and a mean squared residual of 70.18432. To further assess the
model's predictive performance, a scatter plot of actual vs predicted
values is created. This visualization allows for a direct comparison
between the model's predictions and the true-life expectancy values.

```{r}
# Create a scatter plot of actual vs. predicted values for Random Forest model
ggplot(data = test_data, aes(x = LifeExp, y = predictions_rf)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(x = "Actual Life Expectancy", y = "Predicted Life Expectancy", 
       title = "Actual vs. Predicted Life Expectancy (Random Forest Model)") +
  theme_minimal()

```

The scatter plot showcases individual data points where the x-axis
represents the actual life expectancy values, and the y-axis represents
the corresponding predicted values generated by the Random Forest model.
A red dashed line is overlaid on the plot, indicating the ideal scenario
where actual and predicted values perfectly align.

In this case, the scatter plot reveals a generally linear pattern,
suggesting that the Random Forest model captures the underlying
relationships in the data. Points are closely clustered around the ideal
line, indicating a strong correspondence between the predicted and
actual life expectancy values. This alignment signifies that the model
is making accurate predictions, and the deviations from the ideal line
are relatively small.

Overall, the scatter plot visually reinforces the model's effectiveness,
showcasing its ability to provide reliable predictions for life
expectancy based on GDP per capita. The proximity of the points to the
ideal line indicates a strong predictive relationship, supporting the
model's utility in understanding and forecasting life expectancy
patterns.

I will leverage the Support Vector Machine and Gradient Boosting
algorithms to compare their performance with the Linear Regression and
Random Forest models. The objective is to discern which model yields the
most accurate predictions in the context of the relationship between
life expectancy and GDP per capita. By systematically evaluating these
diverse algorithms, I aim to determine the most effective approach for
capturing the intricate relationship and making reliable predictions.

# Build a Support Vector Machine Regressor model

A Support Vector Machine is a supervised machine learning algorithm used
for classification and regression tasks. The main idea behind SVM is to
find a hyperplane that best separates the data into different classes.
The hyperplane chosen is the one that maximizes the margin, which is the
distance between the hyperplane and the nearest data point from each
class.

The equation of a linear Support Vector Machine is as follows: \$\$

f(\mathbf{x}) = \text{sign}(\mathbf{w} \cdot \mathbf{x} + b) \\ \$\$

\$\$ \text{Here:}\\ f(\mathbf{x}) \text{ is the decision function}\\

\mathbf{w} \text{ is the weight vector}\\

\mathbf{x} \text{ is the input vector} \\

b \text{ is the bias term.} \\

$$
The optimization problem associated with SVM can be represented as:
$$

\text{Minimize } \frac{1}{2}\|\mathbf{w}\|\^2\\\
\$\$

\$\$ Subject to the constraints:

y_i(\mathbf{w} \cdot \mathbf{x}\_i + b) \geq 1 \text{ for } i = 1, 2,
\ldots, N \$\$

```{r}
# Load the necessary library
install.packages("e1071")
library(e1071)

# Build a Support Vector Machine Regressor model
svm_model <- svm(LifeExp ~ gdp_per_capita, data = train_data)

# Make predictions on the test set
test_data$predictions_svm <- predict(svm_model, newdata = test_data)

# Calculate R-squared score for SVM
svm_r_squared <- 1 - sum((test_data$LifeExp - test_data$predictions_svm)^2) / sum((test_data$LifeExp - mean(test_data$LifeExp))^2)

print(paste("SVM R-squared Score: ", round(svm_r_squared, 2)))

```

# Build a Gradient Boosting Regressor model

Gradient Boosting is an ensemble learning technique used for both
regression and classification tasks. The algorithm builds a series of
weak learners, typically decision trees, and combines them to create a
strong learner. Gradient Boosting focuses on minimizing the errors of
the previous models by adding new models that correct the mistakes of
the existing ones.

\$\$ F_m(x) = F\_{m-1}(x) + \nu h_m(x)

\$\$

\$\$ \text{where:} \\

F_m(x) \text{ is the model at iteration } m \\

F\_{m-1}(x) \text{ is the model from the previous iteration} \\

\nu \text{ is the learning rate, a hyperparameter in the range (0, 1)}
\\

h_m(x)
\text{ is the weak learner (e.g., a decision tree) trained to correct the residuals}

\$\$

```{r}
# Load the necessary library
install.packages("gbm")
library(gbm)

# Build a Gradient Boosting Regressor model
gbm_model <- gbm(LifeExp ~ gdp_per_capita, data = train_data, n.trees = 100, interaction.depth = 3)

# Make predictions on the test set
test_data$predictions_gbm <- predict(gbm_model, newdata = test_data, n.trees = 100)

# Calculate R-squared score for Gradient Boosting
gbm_r_squared <- 1 - sum((test_data$LifeExp - test_data$predictions_gbm)^2) / sum((test_data$LifeExp - mean(test_data$LifeExp))^2)

print(paste("Gradient Boosting R-squared Score: ", round(gbm_r_squared, 2)))

```

# Model Evaluation

```{r}
# Calculate RMSE for linear regression model
lm_rmse <- sqrt(mean((test_data$LifeExp - predict(lm_model, newdata = test_data))^2))

# Calculate RMSE for Random Forest
rf_rmse <- sqrt(mean((test_data$LifeExp - test_data$predictions_rf)^2))

# Calculate RMSE for Gradient Boosting
gbm_rmse <- sqrt(mean((test_data$LifeExp - test_data$predictions_gbm)^2))

# Calculate RMSE for SVM
svm_rmse <- sqrt(mean((test_data$LifeExp - test_data$predictions_svm)^2))

# Print RMSE values
print(paste("Linear Regression RMSE: ", round(lm_rmse, 2)))
print(paste("Random Forest RMSE: ", round(rf_rmse, 2)))
print(paste("Gradient Boosting RMSE: ", round(gbm_rmse, 2)))
print(paste("SVM RMSE: ", round(svm_rmse, 2)))

```

# RMSE Comparison of Regression Models

Root Mean Squared Error is a metric used to evaluate the accuracy of a
regression model, it provides a measure of how well the model's
predictions match the actual observed values. RMSE measures the average
magnitude of the errors between predicted and actual values and it
penalizes larger errors more heavily than smaller ones.

```{r}

# Create a bar chart to compare RMSE values
model_names <- c("Linear Regression", "Random Forest", "Gradient Boosting", "SVM")
rmse_values <- c(lm_rmse, rf_rmse, gbm_rmse, svm_rmse)

barplot(rmse_values, names.arg = model_names, col = "skyblue", 
        main = "RMSE Comparison of Regression Models",
        ylab = "Root Mean Squared Error (RMSE)", ylim = c(0, max(rmse_values) + 5))

```

Usually, a lower RMSE value indicates better model performance, as it
means the model's predictions are closer to the actual values.
Therefore, based on the above results, the Gradient Boosting model seems
to perform the best among the models, followed closely by the SVM model,
then the Random Forest, and finally Linear Regression.

# R-Squared Comparison of Regression Models

The R-squared is a statistical measure of how well the independent
variables in a regression model explain the variability of the dependent
variable. It is a scale from 0 to 1. where: 0 indicates that the model
does not explain any of the variability in the dependent variable. 1
indicates that the model explains all of the variability in the
dependent variable.

```{r}
r2_lm <- 1 - (sum((test_data$LifeExp - test_data$predictions)^2) / sum((test_data$LifeExp - mean(test_data$LifeExp))^2))
r2_rf <- 1 - (sum((test_data$LifeExp - test_data$predictions_rf)^2) / sum((test_data$LifeExp - mean(test_data$LifeExp))^2))
r2_gbm <- 1 - (sum((test_data$LifeExp - test_data$predictions)^2) / sum((test_data$LifeExp - mean(test_data$LifeExp))^2))
r2_svm <- 1 - (sum((test_data$LifeExp - test_data$predictions_rf)^2) / sum((test_data$LifeExp - mean(test_data$LifeExp))^2))


```

```{r}
# Create a bar chart to compare R-squared values
model_names <- c("Linear Regression", "Random Forest", "Gradient Boosting", "SVM")
r_squared_values <- c(lm_r_squared, rf_r_squared, gbm_r_squared, svm_r_squared)

barplot(r_squared_values, names.arg = model_names, col = "lightgreen", 
        main = "R-Squared Comparison of Regression Models",
        ylab = "R-Squared", ylim = c(0, max(r_squared_values) + 0.2))

# Add values on top of the bars
text(1:4, r_squared_values + 0.05, round(r_squared_values, 2), pos = 3, cex = 1.3, col = "black")

```

R-squared value of 0.68 indicates that the Gradient Boosting model is
reasonably effective at explaining the variation in the target variable.
However, it's crucial to consider other aspects of model evaluation and
the specific goals of this project.

# Limitations:

The project relies on available data, and any gaps or inaccuracies in
the data could impact the results. The models used make certain
assumptions, and the real-world relationship is likely influenced by
numerous complex factors. While the project explores correlations,
establishing causation requires more in-depth studies and considerations
of confounding variables.

# Areas For Further Analysis

To conduct a more extensive analysis, someone can explore how the
relationship evolves over time by incorporating data from additional
years. For a detailed regional analysis, it would be beneficial to
conduct a more granular examination of specific regions, in order to
uncover regional variations and trends. Additionally, investigating
supplementary factors such as healthcare infrastructure, education, and
social policies could provide insights into their contributions to life
expectancy.

# Conculation

This project has undertaken a thorough investigation into the complex
interplay between life expectancy and GDP per capita, leveraging a
diverse set of machine learning models and algorithms. The analysis
unearthed intricate patterns and complexities in the relationship
between these variables. Notably, our key findings illuminate the
logarithmic nature of this relationship, revealing that the influence of
GDP per capita on life expectancy exhibits variations across different
economic contexts.

The results of this project provide valuable insights into understanding
how economic prosperity and life expectancy are linked globally. By
acknowledging and embracing the subtle complexities in these
connections, my research lays the groundwork for more informed and
tailored strategies. Further analyses and careful considerations will
undoubtedly refine our understanding, guiding the development of
comprehensive approaches to tackle urgent challenges in public health
and socio-economic domains.

```{r}


```
